Trump vs. Harris: A Data-Driven Forecast modeling the 2024 US Presidential Elections

### 2024-11-02 paper/paper.qmd
Updated abstract, Introduction and reference

---
title: "Trump vs. Harris: A Data-Driven Forecast modeling the 2024 US Presidential Elections"
subtitle: "A close race but Trump, trumps Harris"
author: 
  - Sophia Brothers
  - Deyi Kong
  - Rayan Awad Alim
thanks: "Code and data are available at: [https://github.com/eeeee-cmd/US_Election/](https://github.com/eeeee-cmd/US_Election/)."
date: today
date-format: long
abstract: "This study models the 2024 U.S. Presidential election outcomes using advanced statistical techniques and predictive modeling. The analysis incorporates data from various polling sources and considers key demographic and political factors to forecast both the popular vote and Electoral College results. The findings suggest a close race between the two main candidates, Donald Trump and Kamala Harris, with a predicted victory for Trump in the Electoral College. This analysis contributes to understanding how public opinion data and electoral systems interact, providing insights into electoral dynamics in contemporary U.S. politics."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Elections in the United States, particularly presidential elections, are pivotal moments that shape not only the nation’s political landscape but also its international relations. The 2024 U.S. Presidential election, like its predecessors, promises to be contentious and closely watched. As electoral outcomes have become increasingly difficult to predict due to polarized political climates and fluctuating public opinion, sophisticated statistical models are essential tools for forecasting results and understanding the dynamics at play.

The core estimand of this study is the projected electoral outcome in terms of state-level victories and Electoral College votes for each candidate. The model leverages polling averages, historical voting patterns, and demographic predictors to estimate the probability of each candidate winning in individual states. The objective is to translate these probabilities into a forecast of Electoral College votes, thus determining the likely overall winner of the election. Given the significant influence of swing states, the model focuses on capturing their dynamics more accurately than national polling averages typically allow.

The results suggest a close race, with Trump holding a slight edge in the Electoral College due to projected victories in key swing states, despite Harris maintaining a lead in national popular vote estimates. Trump's path to victory hinges on critical wins in states such as Pennsylvania, Georgia, and Wisconsin, where his projected margins are narrow but statistically significant. Meanwhile, Harris secures strong performances in populous Democratic strongholds like California and New York, but these do not offset Trump’s advantage in less populous yet crucial battleground states.

Understanding these projected outcomes is essential for several reasons. First, the model highlights the limitations of relying solely on national polling to predict election results, emphasizing the continued relevance of the Electoral College system in determining the presidency. Additionally, it sheds light on the evolving political landscape in swing states, where demographic changes and shifting voter allegiances can have outsized impacts. These insights contribute to a broader understanding of how statistical modeling can inform political strategy and election analysis.

The structure of the paper is as follows: In @sec-data, we provide outlines the data sources and variables considered, followed by the model setup in [@sec-mod] and justification. The results [@sec-result] presents the key findings of the analysis, with a discussion on the implications of these results for political forecasting and public opinion analysis. Finally, the paper concludes with an overview of potential limitations and suggestions for future research in the discussion [@sec-discussion].

@@

The data used in this analysis comes from a combination of publicly available polling data for the 2024 U.S. Presidential election[@FiveThirtyEightPolls]. The analysis leverages the statistical programming language R [@citeR] and several libraries, including `tidyverse` [@tidyverse], `janitor` [@janitor], `knitr` [@knitr], `dplyr` [@dplyr], `purrr` [@purrr], and `here` [@here], for data manipulation, and `ggplot2` [@ggplot] for visualization. The dataset covers various polls conducted across multiple states, capturing the support for each major candidate—Donald Trump and Kamala Harris—along with detailed attributes of the polls.

@@

The primary measurements in the dataset include polling scores, sample sizes, and predicted percentages for each candidate. The data is filtered to include only polls relevant to the two primary candidates, and features like PollRecency were constructed to weigh recent polls higher. The data was cleaned by removing irrelevant or incomplete observations, and new variables were constructed where necessary.

@@

# Model {#sec-mod}

@@

## Model Set-Up

@@

## Model Justification

@@

# Appendix

>>>>

---
title: "Trump vs. Harris: A Data-Driven Forecast modeling the 2024 US Presidential Elections"
subtitle: "A close race but Trump, trumps Harris"
author: 
  - Sophia Brothers
  - Deyi Kong
  - Rayan Awad Alim
thanks: "Code and data are available at: [https://github.com/eeeee-cmd/US_Election/](https://github.com/eeeee-cmd/US_Election/)."
date: today
date-format: long
abstract: "This paper predicts the 2024 U.S. Presidential election, utilizing polling data and demographic factors to predict both popular vote and Electoral College outcomes. The findings indicate a close race between Donald Trump and Kamala Harris, and we predict Trump secures the Electoral College due to projected victories in key swing states while Harris maintains a popular vote lead. The results emphasize the role of state-level dynamics, particularly in swing states, in determining election outcomes. The analysis contributes how public opinion data and electoral systems interact, providing insights into electoral dynamics in contemporary U.S. politics."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

Presidential elections in the United States are pivotal moments that shape not only the nation’s political landscape but also its far-reaching consequences on a global scale. Republicans and Democrats held primary elections across the country earlier in 2024 that eventually filtered down to the main candidates: Donald Trump and Kamala Harris. As electoral outcomes have become increasingly difficult to predict due to polarized political climates and fluctuating public opinion, sophisticated statistical models are essential for forecasting results and understanding the dynamics at play.

The polarized political climate and rapidly shifting public opinion have made predicting outcomes increasingly challenging, where winning key swing states is crucial to securing victory. The core estimand of this study is the projected electoral outcome in terms of state-level victories and Electoral College votes for the candidate. The model incorporates recent polling data, historical voting patterns, and demographic predictors to estimate the probability of each candidate winning in individual states. The objective is to translate these probabilities into a forecast of Electoral College votes, thus determining the likely overall winner of the election.

Our findings reveal a close race, with Trump holding a slight edge in the Electoral College due to projected victories in key swing states, despite Harris maintaining a lead in national popular vote estimates. Trump's path to victory hinges on critical wins in states such as Pennsylvania, Georgia, and Wisconsin, where his projected margins are narrow but statistically significant. Meanwhile, Harris secures strong performances in populous Democratic strongholds like California and New York, but these do not offset Trump’s advantage in less populous yet crucial battleground states. This result underscores the significant influence of swing states and the Electoral College system in determining the outcome, particularly in closely contested elections. These insights contribute to a broader understanding of how statistical modeling can inform political strategy and election analysis.

The structure of the paper is as follows: [@sec-data] outlines the data sources and variables considered, followed by the model setup in [@sec-modset] and justification in [@sec-modjust]. The results in [@sec-result] presents the key findings of the analysis, with a discussion on the implications. [@sec-discussion] then discusses potential limitations and suggestions for future research. [@sec-appx] provides additional detailed information about the data, model and methodology.

@@

The data used in this analysis comes from a combination of publicly available polling data for the 2024 U.S. Presidential election [@FiveThirtyEightPolls]. The analysis leverages the statistical programming language R [@citeR] and several libraries, including `tidyverse` [@tidyverse], `janitor` [@janitor], `knitr` [@knitr], `dplyr` [@dplyr], `arrow` [@arrow], `purrr` [@purrr], and `here` [@here], for data manipulation. `ggplot2` [@ggplot] and `kableExtra` [@kableExtra] for visualization. The dataset covers various polls conducted across multiple states, capturing the support for each major candidate—Donald Trump and Kamala Harris—along with detailed attributes of the polls.

@@

The primary measurements in the dataset include polling scores, sample sizes, and predicted percentages for each candidate. The data is filtered to include only polls relevant to the two primary candidates，Trump and Harris, and features like PollRecency were constructed to weigh recent polls higher. The data was cleaned by removing irrelevant or incomplete observations, and new variables were constructed where necessary.

@@

# Model

@@

## Model Set-Up {#sec-modset}

@@

## Model Justification {#sec-modjust}

@@

# Appendix {#sec-appx}
